Towards a standardized evaluation of imputation methodology

### 2022-09-21 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
Fix typos/referencing errors, combine paragraphs 1 and 2 from section 3

A well-designed simulation does not guarantee fair evaluations. Estimands and other simulation targets should be clearly and unambiguously defined in the context of the study aim \citet{pete14}. The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error; see][]{morr18}. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods--i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions \citep{scho18}. If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons.

@@

An obvious problem with design-based simulation is that obtaining a large dataset without missing entries can be challenging. Most real-world data contains at least some missing entries, for which the true underlying missing data model is--by definition--unknown. Therefore, the simulator needs to deal with missingness in \textit{some} way before incomplete empirical data can serve as comparative truth in the simulations. It may seem like an intuitive solution to only draw complete cases from the large dataset, which would indeed yield complete samples. However, there may be inherent differences between cases with and cases without any missing values, due to the unknown missing data model. Only sampling complete cases from the data set may thus result in samples that fail to capture all relevant real-world conditions, which in turn refutes the main reason for using design-based simulation. Another way to deal with missingness in a design-based simulation is to impute the incomplete dataset once, to obtain a single completed dataset to draw samples from. Unfortunately, this practice may favor the imputation method that was used in this initial imputation step throughout any further evaluations. Just to be clear: leaving the missingness as-is and inducing additional missing values to impute is no option here, because we would not have a real and unbiased comparative truth.

@@

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term `missingness proportion' restricted to the variable-by-variable interpretation. The distinction between the two concepts, however, is not always clear in the literature, and convoluting the terms may lead to incorrect generalizations because they rarely mean the same (e.g., a proportion of 50\% incomplete cases in bivariate data could translate to a missingness proportion of 25\% in both variables, or one completely observed variable and one with 50\% missingness). Simulators should therefore be explicit in their description of the missing data generating model and should make sure that the model fits the aim of their simulations.

Extrapolating simulation results from a specific missing data generating model to more intricate (empirical) missingness could lead to sub-optimal advice in practice. For example, if a simulator induces missingness in the outcome variable of their analysis model exclusively, they may inadvertedly induce missingness according to the special case descibed by \citet[][p. \S 2.7]{buur18}. Under this specific missingness model, list-wise deletion may outperform any imputation method. Such a conclusion would, unfortunately, only translate to data that adhere to the same special case, while biasing inferences in other cases. % Another example of a missingness pattern that may inadvertently impose assumptions on generalizations is a monotone pattern \citep[i.e. a pattern with uniformly increasinmissingness proportions along variables;][]{litt20}. If a simulator investigates iterative imputation methods under monotone missingness, they might believe that the evaluated imputation algorithms converge instantly, while these methods actually require iteration to produce valid results in all other situations. 
Another example of potential side-effects of missingness generation may occur when multivariate missingness is generated using step-wise univariate missingness induction. The resulting incomplete data may then not have the desired statistical properties \citep{ampute}. This complicates any definitive conclusions about imputation method performance in relation to the data generation parameters.

@@

Besides MCAR missingness, observed-data dependent missingness mechanisms--such as missing at random (MAR) missingness--are paramount in statistical inference investigations. A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data, generating the missingness through autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly}. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}.

@@

In addition to varying the missingness mechanisms, each simulated mechanism should be combined with different missingness patterns. Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process \citep[Corollary 6.1A]{litt20}. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter-- the inference does not depend on the assumption.

Alternatively, the validity of assumptions becomes increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under several missingness proportions to amulate a realistic range in severity of the missingness problem. Depending on how the missingness mechanism interacts with the simulated data, higher missingness proportions may yield biased or invalid completed data inferences. The missingness proportions should therefore be considered carefully.

@@

At minimum, imputation method performance should be quantified using appropriate measures. It depends on the specifics of each simulation study which performance measures would be most suitable. We recommend to evaluate at least the following points:

(i) The methods should preferably be unbiased. Note that the way bias is calculated should be carefully chosen and described, since this can greatly influence the interpretation of the results. A negligible absolute bias, for example, may already yield an almost infinite relative bias if the true value of the estimand is zero. In most cases, unbiased estimates may be expected under a MCAR missingness mechanism which can be easily verified by including MCAR missingness to the scope of missingness investigations.

(ii) The intervals for estimates should have a valid coverage of the population (i.e. true) value. Coverage of a 95\% interval should in theory be $\geq 95$, where a coverage rate of 95\% would be most efficient \citep[p. 589]{neym34}. Under-coverage indicates that the procedure is not confidence valid and may lead to invalid inference. Under-coverage may occur when the estimation procedure is either biased, too liberal leading to intervals that are too narrow, or both. Over-coverage, though technically not confidence invalid, would indicate that efficiency could still be gained. In such cases, a narrower interval exists for which the method would still yield confidence valid result.

@@

(iv) Predictive accuracy may be quantified using the root mean squared error (RMSE). We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values and it may increase the type I error rate of the statistical inferences \citep[Chapter 2.6][]{buur18}. However, if a study is aimed at obtaining predictions, and inferential validity is not of interest, the RMSE of the predicted values can be compared between method, which may yield valuable information about the methods' efficiency in terms of both accuracy and precision.

If the goal is inference--not prediction-- then the uncertainty about estimates needs to be properly quantified. To capture this uncertainty, the standard errors of the estimates should be correctly calculated, which requires multiple imputation. The aim of multiple imputation is not to reproduce the data, but to allow for obtaining valid inference given that the data are incomplete. This means that, given the framework provided by \citet{rubi87}, statistical properties such as bias, confidence intervals, and the coverage rate of the confidence intervals should always be studied.

After establishing the statistical validity of an imputation routine on the level of the analysis estimates, it remains paramount to also study the imputation-generating process. Omitting such investigations may yield sub-optimal results. One simple example arises often in practice: generating implausible imputed values. Even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny. Under many circumstances, imputation methods may be realistically expected to preserve both marginal and conditional distributions with respect to the comparative truth. Imputation methods that fail to do so, should not be considered general-purpose methods. Many techniques can yield valid inferences, but the techniques that samples realistic or plausible values may be preferable in practice.

@@

After the careful consideration and execution of all simulation aspects, the evaluations should be properly reported. The inconsistent display of simulation conditions may impact the objectivity of meta-evaluations over imputation methods, as one method's performance may appear to be favorable because of less stringent simulation conditions. This ultimately may lead to statisticians recommending a less efficient method to applied researchers, thereby limiting the efficiency of the imputation approach and unnecessarily lowering the statistical power.

We encourage simulators to document their choices and to be explicit in their descriptions. For example, the simulation design may be presented textually, in a flow chart or as a block of pseudo-code, whereas missingness mechanisms may be written as a function of the data or displayed graphically. Ideally, this should be supplemented by an online repository with all of the data and code required to reproduce the simulation results. To aid simulators in reporting and move towards standardization in evaluation, we provide a draft version for reporting guidelines in Appendix A.1 (also available from \underline{www.gerkovink.com/evaluation}). We invite the readers of this paper to contribute to its development.

>>>>

A well-designed simulation does not guarantee fair evaluations. Estimands and other simulation targets should be clearly and unambiguously defined in the context of the study aim \citep{pete14}. The study aim and the required level of precision may also inform the number of simulation repetitions \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error; see][]{morr18}. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods--i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions \citep{scho18}. If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons.

@@

An obvious problem with design-based simulation is that obtaining a large dataset without missing entries can be challenging. Most real-world data contains at least some missing entries, for which the true underlying missing data model is--by definition--unknown. Therefore, the simulator needs to deal with missingness in \textit{some} way before incomplete empirical data can serve as comparative truth in the simulations. It might seem like an intuitive solution to only draw complete cases from the large dataset, which would indeed yield complete samples. However, there may be inherent differences between cases with and cases without any missing values, due to the unknown missing data model. Only sampling complete cases from the data set may thus result in samples that fail to capture all relevant real-world conditions, which in turn refutes the main reason for using design-based simulation. Another way to deal with missingness in a design-based simulation is to impute the incomplete dataset once, to obtain a single completed dataset to draw samples from. Unfortunately, this practice may favor the imputation method that was used in this initial imputation step throughout any further evaluations. Just to be clear: leaving the missingness as-is and inducing additional missing values to impute is no option here, because we would not have a real and unbiased comparative truth.

@@

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term `missingness proportion' restricted to the variable-by-variable interpretation. The distinction between the two concepts, however, is not always clear in the literature, and convoluting the terms may lead to incorrect generalizations, because they rarely mean the same (e.g., a proportion of 50\% incomplete cases in bivariate data could translate to a missingness proportion of 25\% in both variables, or one completely observed variable and one with 50\% missingness). Simulators should therefore be explicit in their description of the missing data generating model and should make sure that the model fits the aim of their simulations.

Extrapolating simulation results from a specific missing data generating model to more intricate (empirical) missingness could lead to sub-optimal advice in practice. For example, if a simulator induces missingness in the outcome variable of their analysis model exclusively, they may inadvertedly induce missingness according to the special case descibed by \citet[][\S 2.7]{buur18}. Under this specific missingness model, list-wise deletion may outperform any imputation method. Such a conclusion would, unfortunately, only translate to data that adhere to the same special case, while biasing inferences in other cases. % Another example of a missingness pattern that may inadvertently impose assumptions on generalizations is a monotone pattern \citep[i.e. a pattern with uniformly increasinmissingness proportions along variables;][]{litt20}. If a simulator investigates iterative imputation methods under monotone missingness, they might believe that the evaluated imputation algorithms converge instantly, while these methods actually require iteration to produce valid results in all other situations. 
Another example of potential side-effects of missingness generation may occur when multivariate missingness is generated using step-wise univariate missingness induction. The resulting incomplete data may then not have the desired statistical properties \citep{ampute}. This complicates any definitive conclusions about imputation method performance in relation to the data generation parameters.

@@

Besides MCAR missingness, observed-data dependent missingness mechanisms--such as missing at random (MAR) missingness--are paramount in statistical inference investigations. A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}; generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data, generating the missingness through autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly}. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}.

@@

In addition to varying the missingness mechanisms, each simulated mechanism should be combined with different missingness patterns. Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process \citep[Corollary 6.1A]{litt20}. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter-- the inference does not depend on the assumption. Alternatively, the validity of assumptions becomes increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under several missingness proportions to amulate a realistic range in severity of the missingness problem. Depending on how the missingness mechanism interacts with the simulated data, higher missingness proportions may yield biased or invalid completed data inferences. The missingness proportions should thus be considered carefully.

@@

At minimum, imputation method performance should be quantified using appropriate measures. It depends on the specifics of each simulation study which performance measures would be most suitable. If the goal is inference--not prediction--the uncertainty about estimates needs to be properly quantified. To capture this uncertainty, the standard errors of the estimates should be correctly calculated, which requires multiple imputation. The aim of multiple imputation is not to reproduce the data, but to allow for obtaining valid inference given that the data are incomplete. This means that, given the framework provided by \citet{rubi87}, statistical properties such as bias, confidence intervals, and the coverage rate of the confidence intervals should always be studied. We therefore recommend to evaluate the following points:

(i) The methods should preferably be unbiased. Note that the way bias is calculated should be carefully chosen and described, since this can greatly influence the interpretation of the results. A negligible absolute bias, for example, may already yield an almost infinite relative bias if the true value of the estimand is zero. In most cases, unbiased estimates may be expected under a MCAR missingness mechanism, which can be easily verified by including MCAR missingness to the scope of missingness investigations.

(ii) The intervals for estimates should have a valid coverage of the population (i.e. true) value. Coverage of a 95\% interval should in theory be $\geq 95$\%, where a coverage rate of 95\% would be most efficient \citep[p. 589]{neym34}. Under-coverage indicates that the procedure is not confidence valid and may lead to invalid inference. Under-coverage may occur when the estimation procedure is either biased, too liberal leading to intervals that are too narrow, or both. Over-coverage, though technically not confidence invalid, would indicate that efficiency could still be gained. In such cases, a narrower interval exists for which the method would still yield confidence valid result.

@@

(iv) Predictive accuracy may be quantified using the root mean squared error (RMSE). We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values and it may inflate the type I error rate of statistical inferences \citep[Chapter 2.6][]{buur18}. However, if a study is aimed at obtaining predictions, and inferential validity is not of interest, the RMSE of the predicted values can be compared between methods. Then, the RMSE may yield valuable information about the methods' comparative efficiency in terms of accuracy and precision.

After establishing the statistical validity of an imputation routine on the level of the analysis estimates, it remains paramount to also study the imputation-generating process. Omitting such investigations may yield sub-optimal results. One simple example arises often in practice: generating implausible imputed values. Even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny. Under many circumstances, imputation methods may be realistically expected to preserve both marginal and conditional distributions with respect to the comparative truth. Imputation methods that fail to do so, should not be considered general-purpose methods. Many techniques can yield valid inferences, but techniques that sample realistic or plausible values may be preferable in practice.

@@

We encourage simulators to carefully consider and document their choices in the evaluation of imputation methodology. Simulation studies should not only be well-executed, but also properly reported. For example, the inconsistent display of simulation conditions may impact the objectivity of meta-evaluations over imputation methods, as one method's performance may appear to be favorable because of less stringent simulation conditions. This may ultimately lead to statisticians recommending a less efficient method to applied researchers, thereby limiting the efficiency of the imputation approach and unnecessarily lowering the statistical power. Simulators should therefore to be explicit in the descriptions of their evaluations, but deciding what and how to report may be challenging. The simulation design could be presented textually, in a flow chart or as a block of pseudo-code, whereas missingness mechanisms could be written as a function of the data or displayed graphically. Ideally, the evaluations should be supplemented by an online repository with all of the data and code required to reproduce the simulation results. To aid simulators in reporting and move towards standardization in evaluation, we provide a draft version for reporting guidelines in Appendix A.1 (also available from \underline{www.gerkovink.com/evaluation}). We invite the readers of this paper to contribute to its development.
